### 2018-04-05 ident/docs/draft_1.tex
significant changes to intro based on prof. M's comment

\begin{document}
	\begin{center}
		\begin{Large}
			Practical Identification and Experimental Design for Parameter Estimation in Kinetic Models of Metabolism
		\end{Large}\\
		Shyam Srinivasan\textsuperscript{a}, William R. Cluett\textsuperscript{a} and Radhakrishnan Mahadevan\textsuperscript{*,a,b}\\
	\end{center}
	a - Department of Chemical Engineering and Applied Chemistry, University of Toronto, Toronto, ON, Canada.\\
	b - Institute for Biomaterials and Biomedical Engineering, University of Toronto, Toronto, ON, Canada.\\
	{*} Corresponding author
	\section*{Abstract}
	\section{Introduction:}
	Kinetic models of metabolism can be used to study the dynamic characteristics of metabolic networks \parencite{Bordbar2015, Andreozzi2016}.

	Unlike CBMs, where the ability to predict the steady state responses of metabolic networks is dependent on the stoichiometry of the network, the prediction of dynamic responses of metabolic networks to different perturbations using kinetic models of metabolism is also dependent on the numerical values of the enzyme kinetic parameters ($\theta$ in Equation \ref{eq:kinstoich}) \parencite{Andreozzi2016a}. Analyzing the ability of a metabolic network to exhibit dynamic characteristics like multiple steady states and oscillations, irrespective of the structure of the network, is one example where parameter values might play a crucial role \parencite{Srinivasan2015, Vital-Lopez2006}. The use of in vitro, or unreliable in vivo parameter estimates reduces confidence in the model predicted behaviour \parencite{Andreozzi2016a}. Consequently, the reduction in confidence hampers the use of these models to gain insight into the functioning of metabolic networks \parencite{Chakrabarti2013a, Bordbar2015}. The corresponding increase in uncertainty in model predicted responses becomes an obstacle for using the predicted responses as a basis for designing the metabolic networks to achieve any of the aforementioned goals of metabolic engineering and design.

@@

	\begin{figure}[!tbhp]
		\centering{\includegraphics[width=1.0\textwidth,height=1.0\textheight,keepaspectratio]{figures/figure4/v1_v1max_v2_v3_1_data_utility}}
		\caption{Utility of experimental data combinations on the basis of their ability to identify the most number of parameters. Information is shown for parameters modeling fluxes a) $v_1$, b) $v_2$ and c) $v_3$. The total number of combinations of experimental data is shown in the vertical axis and the horizontal axis shows the total possible number of parameters that can be identified by data from combinations of a) two, b) two, and c) three experiments. The percentages shown in the plots represent the fraction of the total combinations used to test identifiability of parameters for a given flux. A total of 240 data combinations are used for identifiability analysis for a) $v_1$ and b) $v_2$, and 3360 combinations are used to practically identify c) $v_3$. Section \ref{sec:experiments} provides more details on how the combinations of experimental data are generated.}\label{fig:figure4}
	\end{figure}

	\begin{figure}[!tbhp]
		\centering{\includegraphics[width=1.0\textwidth,height=1.0\textheight,keepaspectratio]{figures/figure1/v1_k1cat_v1max_v2_v3_1_2_ident}}
		\caption{The number of data combination from 18 different in silico experiments that can practically identify each parameter in fluxes a) $v_1$, b) $v_2$ and c) $v_3$ when there is no noise in the input experimental data. The percentage of total experimental data combinations that can identify each parameter is specified either in, or right next to the bar showing the total number of combinations identifying a given parameter. Since the total number of experiments used to identify parameters in $v_1$ and $v_2$ is two, the total number of data combinations used to identify parameters in fluxes $v_1$ and $v_2$ is lower (at 306) that the total number of combinations used to test identifiability of parameters for $v_3$ (at 4896), which is modeled with three parameters and requires data from at least three different experiments.}\label{fig:figure1}
	\end{figure}

	\begin{figure}[!tbhp]
		\centering{\includegraphics[width=1.0\textwidth,height=0.6\textheight, keepaspectratio]{figures/figure2/v1_v1max_v2_exp_info}}
		\caption{The contribution of different experiments types used in a combination of two experiments ($k = \{1, 2\}$) that can practically identify parameter a) $V_1^{max}$, b) $K_1^{ac}$, c) $V_2^{max}$ and d) $K_2^{pep}$. The percentages reflect the fractional contribution of each experiment type towards all identifiable data combinations.}\label{fig:figure2}
	\end{figure}

>>>>

\begin{document}
	\begin{center}
		\begin{Large}
			Practical Identification and Experimental Design for Parameter Estimation in Kinetic Models of Metabolism
		\end{Large}\\
		Shyam Srinivasan\textsuperscript{a}, William R. Cluett\textsuperscript{a} and Radhakrishnan Mahadevan\textsuperscript{*,a,b}\\
	\end{center}
	a - Department of Chemical Engineering and Applied Chemistry, University of Toronto, Toronto, ON, Canada.\\
	b - Institute for Biomaterials and Biomedical Engineering, University of Toronto, Toronto, ON, Canada.\\
	{*} Corresponding author
	\section*{Abstract}
	\section{Introduction:}
	The use of metabolic engineering spans a wide variety of applications, ranging from the design of microorganisms for the biosynthesis of commodity and specialty chemicals \parencite{Andreozzi2016} to engineering mammalian cells as therapeutic targets for cures to various ailments that affect humans \parencite{Bordbar2015,DiFilippo2016,Apaolaza2017}. These applications require us to understand the numerous complex interaction, their roles in cell function, and sometimes even the mechanisms behind these interactions. Computational models offer an avenue to study and understand these interactions through simplified representations of these interactions. They are also used to predict changes in cell function based on changes in the type and nature of the modeled interactions, or finding therapeutic targets to cure various ailments \textcolor{red}{(cite from old draft)}.

	Constraint-based models (CBMs) of metabolism are one such category of models that are used to improve our understanding of the flow of flux through metabolism by representing it as a stoichiometric network of reactions that operate under a pseudo-steady state assumption. The ability of CBMs to shine light on the nonintuitive interactions that govern cellular metabolism is leveraged to predict the effect of different metabolic engineering design decisions on the ability of a cell to grow, or produce a desired metabolite (Bordbar, Monk, et al. 2014; Maia, M. Rocha, and I. Rocha 2016). These advantages, combined with minimal number of parameters that are required to make CBMs functional, have made them useful to get an initial understanding of metabolism and its relationship to cellular function for various aforementioned applications \textcolor{red}{(CBM use references? cMCS, MCS may be?)}.

	However, since metabolism is assumed to operate under a pseudo steady state, the metabolite concentrations within the metabolic network are assumed to be constant, and CBMs of metabolism do not model changes in metabolite concentrations. Furthermore, since CBMs represent metabolism using only the stoichiometry of its constituent reactions, they do not account for the various non-catalytic regulatory interactions that are also responsible of metabolic function. These shortcomings prevent CBMs from being used to fully understand the steady state as well as the dynamic characteristics of metabolic networks.

	Recently, the focus has shifted towards understanding the importance of the dynamics of metabolism in determining cell function \textcolor{red}{(recent kinetic model review about their utility)}. Consequently, it has been shown that using kinetic models that can represent the dynamics of the network can not only help us learn about previously unknown characteristics of metabolism \textcolor{red}{(red blood cell-both old and new, Maranas bifurcation paper)}, but can also improve predictions about the impact of engineering design perturbations on metabolism \textcolor{red}{(ecoli k457, k-optForce, Maranas kinetic model paper)}. \parencite{Bordbar2015, Andreozzi2016}. These kinetic models of metabolism, in contrast to their constraint-based counterparts, account for the thermodynamic and regulatory interactions that underly metabolic networks in addition to its stoichiometry \parencite{Andreozzi2016a}.

	Kinetic models differ from CBMs is their use of mechanistic and empirical enzyme kinetic rate laws to model enzyme catalyzed fluxes within a metabolic network. The other aspect is the fact that kinetic models are heavily parameterized. These parameters represent various aspects of the enzyme kinetic rate laws \textcolor{red}{(kinetic model reviews, our review and a more recent one)}. Hence, the use of kinetic models requires information on all the enzyme kinetic rate laws that will be used to model all the fluxes within a metabolic network, as well as numerical values for all the parameters used in these rate laws. These aspects not only differentiate kinetic models from CBMs of metabolism, but are also two of the major hurdles to their widespread use and adoption.

	Firstly, the development of these enzyme kinetic rate laws is based on observations of enzyme activity that are made in vitro. Accordingly, several authors have questioned their relevance for gleaning information on the dynamics of metabolism under in vivo conditions, as opposed to in vitro conditions \textcolor{red}{(Heijnen and Klamt)}. This problem is further compounded by the fact that typical parameter values used in kinetic models of metabolism are estimated in vitro, and may not be applicable under in vivo conditions. Any available in vivo parameter estimates are associated with large uncertainties. Analyzing the ability of a metabolic network to exhibit dynamic characteristics like multiple steady states and oscillations, irrespective of the structure of the network, is one example where parameter values might play a crucial role \parencite{Srinivasan2015, Vital-Lopez2006,Srinivasan2017}. Hence, the ability to study the steady state and dynamic responses of metabolic networks, under in vivo conditions, to different perturbations is compromised. The use of in vitro, or unreliable in vivo parameter estimates reduces confidence in the model predicted behaviour \parencite{Andreozzi2016a}. Consequently, the reduction in confidence hampers the use of these models to gain insight into metabolic network function \parencite{Chakrabarti2013a, Bordbar2015}. The corresponding increase in uncertainty in model predicted responses becomes an obstacle for using the predicted responses as a basis for designing the metabolic networks to achieve any of the aforementioned goals of metabolic engineering and design.

	Efforts to improve kinetic models have been dedicated to the development of newer enzyme kinetic models that have a higher correlation with what is observed in vivo. Authors have taken different approaches to address this problem. Some have focused on the development and improvement of parameter estimation methods based on optimization principles for obtaining true in vivo parameter values based on available in vivo experimental data without changing the underlying kinetic rate laws used in the model \textcolor{red}{Isabel Rocha papers on optimization methods, Japanese group that did parameter estimation using a HPC cluster}. Others have sought to completely abandon in vitro kinetic rate law descriptions in favour of approximate kinetic rate law models whose parameters are more informative and can be estimated in a relatively easier fashion \textcolor{red}{(Heijnen and other approximate kinetic modeling formats papers, Kotte 2010)}.

	Despite these efforts, one of persistent problems is the lingering uncertainty in the parameter estimates and the subsequent predictions that are obtained from these methods. Some authors have proposed to alleviate this parameter uncertainty by using a Monte Carlo approach to obtaining kinetic models of metabolism \textcolor{red}{(ensemble modeling and ORACLE)}. Here, experimental steady state data is used as a reference point and kinetic models are built around this steady state value. Although a some have sought to quantify this uncertainty in parameter estimation through methods based on sensitivity analysis \textcolor{red}{(Nathan van Riel and Heijnen papers on uncertainty analysis)}, others have sought to approach parameter estimation from a Bayesian perspective \textcolor{red}{(Lars Nielsen papers)}.

	These new kinetic models are either formulated on the basis of approximate enzyme kinetic rate laws to alleviate

	In the absence of any in vivo parameter estimates, the norm is to either use available in vitro enzyme kinetic parameters, or use different parameter estimation techniques to determine parameter values based on either in vitro or in vivo data. So, the ability of these enzyme rate laws, and kinetic models built with these rate laws to study, understand and predict in vivo metabolic network function is limited \textcolor{red}{Heijnen}.

	 but have also focused on methods to obtain parameter values that are applicable under in vivo conditions, as opposed to in vitro enzyme kinetic parameters. Research in the former area has resulted in the proposal of numerous approximate kinetic models of metabolism

	the prediction of dynamic responses of metabolic networks to different perturbations using kinetic models of metabolism is dependent on numerous enzyme kinetic parameters.  that are used to represent the changes in both concentrations and fluxes

	the numerical values of the enzyme kinetic parameters ($\theta$ in Equation \ref{eq:kinstoich}) \parencite{Andreozzi2016a}. Analyzing the ability of a metabolic network to exhibit dynamic characteristics like multiple steady states and oscillations, irrespective of the structure of the network, is one example where parameter values might play a crucial role \parencite{Srinivasan2015, Vital-Lopez2006}.

@@

	\begin{figure}[!tbhp]
		\centering{\includegraphics[width=1.0\textwidth,height=1.0\textheight,keepaspectratio]{figures/figure4/v1_v1max_v2_v3_1_data_utility}}
		\caption{Utility of experimental data combinations on the basis of their ability to identify the most number of parameters. Information is shown for parameters modeling fluxes a) $v_1$, b) $v_2$ and c) $v_3$. The total number of combinations of experimental data is shown in the vertical axis and the horizontal axis shows the total possible number of parameters that can be identified by data from combinations of a) two, b) two, and c) three experiments. The percentages shown in the plots represent the fraction of the total combinations used to test identifiability of parameters for a given flux. A total of 240 data combinations are used for identifiability analysis for a) $v_1$ and b) $v_2$, and 3360 combinations are used to practically identify c) $v_3$. Section \ref{sec:experiments} provides more details on how the combinations of experimental data are generated.}\label{fig:data_utility}
	\end{figure}

	\begin{figure}[!tbhp]
		\centering{\includegraphics[width=1.0\textwidth,height=1.0\textheight,keepaspectratio]{figures/figure1/v1_k1cat_v1max_v2_v3_1_2_ident}}
		\caption{The number of data combination from 18 different in silico experiments that can practically identify each parameter in fluxes a) $v_1$, b) $v_2$ and c) $v_3$ when there is no noise in the input experimental data. The percentage of total experimental data combinations that can identify each parameter is specified either in, or right next to the bar showing the total number of combinations identifying a given parameter. Since the total number of experiments used to identify parameters in $v_1$ and $v_2$ is two, the total number of data combinations used to identify parameters in fluxes $v_1$ and $v_2$ is lower (at 306) that the total number of combinations used to test identifiability of parameters for $v_3$ (at 4896), which is modeled with three parameters and requires data from at least three different experiments.}\label{fig:ident}
	\end{figure}

	\begin{figure}[!tbhp]
		\centering{\includegraphics[width=1.0\textwidth,height=0.6\textheight, keepaspectratio]{figures/figure2/v1_v1max_v2_exp_info}}
		\caption{The contribution of different experiments types used in a combination of two experiments ($k = \{1, 2\}$) that can practically identify parameter a) $V_1^{max}$, b) $K_1^{ac}$, c) $V_2^{max}$ and d) $K_2^{pep}$. The percentages reflect the fractional contribution of each experiment type towards all identifiable data combinations.}\label{fig:exp_info}
	\end{figure}
